Allee Effect and Cultural Evolution

### 2015-05-04 manuscript/draft.tex
Updated Abstract

\journal{Evolution and Human Behavior}

@@

\author[label1]{Enrico R. Crema\corref{cor1}}
\cortext[cor1]{Enrico R. Crema, enrico.crema@upf.edu}
\author[label2]{Xavier Rubio-Campillo}

\address[label1]{CaSEs - Complexity and Socio-Ecological Dynamics Research Group, Barcelona}
\address[label2]{BSC - Barcelona Supercomputing Center}
\begin{abstract}
Social learning often relies on the evaluation of the fitness of cultural traits through a comparative assessment of alternatives observed in a population. This process is shaped by the details of the learning strategy, as well as the context where the trait is displayed. The latter can modify the fitness of the cultural trait, ultimately changing the evolutionary fate of each adoption or loss. Here we explore a particular situation where the fitness of a trait is dependent to its frequency in the population.  We argue that many cultural traits pertaining subsistence activities exhibit both positive and negative frequency selection, derived by the beneficial effect of cooperation and the detrimental role played by interference and limited resources. A too small or too large population of individuals possessing the same variant will be subject to a decline in fitness, while an intermediate size will be optimal. We explore this core assumption, which closely resembles the “Allee effect” in ecology, using a modified Lotka-Volterra model of payoff-biased social learning, and show that this can generate a large variety of evolutionary dynamics observed in socially transmitted subsistence traits,  including episodes of reversion to suboptimal variants, long-term co-existence of multiple variants, and fixation to a single optimal or suboptimal strategy.
\end{abstract}

>>>>

\journal{Theoretical Population Biology}

@@

\author[label1,label2]{Enrico R. Crema\corref{cor1}}
\cortext[cor1]{Corresponding Author: enrico.crema@upf.edu}
\author[label3]{Xavier Rubio-Campillo}

\address[label1]{CaSEs - Complexity and Socio-Ecological Dynamics Research Group, Barcelona}
\address[label2]{UCL Institute of Archaeology}
\address[label3]{BSC - Barcelona Supercomputing Center}

\begin{abstract}
Several forms of social learning rules are guided by the amount of payoff received by social teachers adopting specific cultural variants. The rationale, is that such payoff will be correlated to the beneficial or detrimental properties of a given variant. Here we assume that the payoff directly affects the reproductive fitness of individuals, and that it is determined by the absolute frequency of individuals possessing the same cultural variant. More specifically we consider a  form of frequency dependence where a too small or too large population of individuals possessing the same variant will be subject to a decline in fitness, while an intermediate size is optimal. This non-linear relationship, known as Allee effect in ecology, portrays a large number of cultural variants, in particular subsistence strategies that are enhanced  by the effect of cooperation but are constrained by interference and limited resources. This paper explores the evolutionary dynamics of a two trait model where each variant exhibit Allee effect. We used a modified Lotka-Volterra model of payoff-biased social learning, and show that that different degrees of reliance on social learning can exhibit a variety of equilibria, including episodes of reversion to suboptimal variants, long-term co-existence of multiple variants, and fixation to a single optimal or suboptimal strategy.
\end{abstract}
